---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

<div id="devex-badge"><a rel="Exploration" href="https://github.com/BCDevExchange/docs/blob/master/discussion/projectstates.md"><img alt="Being designed and built, but in the lab. May change, disappear, or be buggy." style="border-width:0" src="https://assets.bcdevexchange.org/images/badges/exploration.svg" title="Being designed and built, but in the lab. May change, disappear, or be buggy." />
[![Travis-CI Build Status](https://travis-ci.org/bcgov/ssdca.svg?branch=master)](https://travis-ci.org/bcgov/ssdca)
[![Coverage Status](https://img.shields.io/codecov/c/github/bcgov/ssdca/master.svg)](https://codecov.io/github/bcgov/ssdca?branch=master)

# ssdca

`ssdca` is an R package to fit Species Sensitivity Distributions using Maximum Likelihood and plot the data and results.

This package is under development for the B.C. Ministry of Environment by [Poisson Consulting](https://github.com/poissonconsulting). 

## Installation

To install the latest version of `ssdca`:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("bcgov/ssdca")
library(ssdca)
```

## Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/ssdca/issues/).

## How to Contribute

If you would like to contribute to the package, please see our 
[CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). 
By participating in this project you agree to abide by its terms.

## Source Data

## License

The code is released under the Apache License 2.0

    Copyright 2015 Province of British Columbia

    Licensed under the Apache License, Version 2.0 (the "License");
    you may not use this file except in compliance with the License.
    You may obtain a copy of the License at 

       http://www.apache.org/licenses/LICENSE-2.0

    Unless required by applicable law or agreed to in writing, software
    distributed under the License is distributed on an "AS IS" BASIS,
    WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
    See the License for the specific language governing permissions and
    limitations under the License.
    
The data are licensed under the ([Open Government Licence - Canada](http://open.canada.ca/en/open-government-licence-canada)).
See the `data-raw` folder for more information.

