---
title: "Additional technical details"
author: "sstools Team"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
csl: my-style.csl
latex_engine: MathJax
mainfont: Arial
mathfont: Courier
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Additional technical details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>

.box {
  height: auto;
  width: 65%;
  padding: 10px;
  border: 1.5px outset #5C5B5D;
  background-color: #E7E8EC;
  border-radius: 8px;
  text-align: left;
  margin: auto;
}

</style>

<script type="text/x-mathjax-config">MathJax.Hub.Config({  "HTML-CSS": { minScaleAdjust: 0.5, availableFonts: [] }  });
</script>


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  fig.width = 6,
  fig.height = 4
)
```

## Small sample bias

The ssdtools package uses the method of Maximum Likelihood (ML) to estimate parameters for each distribution
that is fit to the data. Statistical theory says that maximum likelihood estimators are asymptotically
unbiased, but does not guarantee performance in small samples. A detailed account of the issue of small sample bias in estimates can be found in the following [pdf](https://github.com/bcgov/ssdtools/blob/master/vignettes/small-sample-bias.pdf).

## The inverse Pareto and inverse Weibull as limiting distributions of the Burr Type-III distribution

### Burr III distribution

The probability density function, ${f_X}(x;b,c,k)$ and cumulative distribution function, ${F_X}(x;b,c,k)$ for the Burr III distribution (also known as the *Dagum* distribution) as used in `ssdtools` are:


<div style="text-align: center;"font-size: 1.2em; color:#3664EE;" class="box";>
<b><u>Burr III distribution</u></b>\[\begin{array}{*{20}{c}}
{{f_X}(x;b,c,k) = \frac{{b\,k\,c}}{{{x^2}}}\frac{{{{\left( {\frac{b}{x}} \right)}^{c - 1}}}}{{{{\left[ {1 + {{\left( {\frac{b}{x}} \right)}^c}} \right]}^{k + 1}}}}{\rm{ }}}&{b,c,k,x > 0}
\end{array}\]<br>\[\begin{array}{*{20}{c}}
{{F_X}(x;b,c,k) = \frac{1}{{{{\left[ {1 + {{\left( {\frac{b}{x}} \right)}^c}} \right]}^k}}}{\rm{ }}}&{b,c,k,x > 0}
\end{array}\]
</div>

### Inverse Pareto distribution

<br>Let $X \sim Burr(b,c,k)$ have the *pdf* given in the box above. It is well known that the distribution of $Y = \frac{1}{X}$ is the *inverse Burr* distribution (also known as the *SinghMaddala* distribution) for which:\[\begin{array}{*{20}{c}}
{{f_Y}(y;b,c,k) = \frac{{c{\kern 1pt} {\kern 1pt} k{{\left( {\frac{y}{b}} \right)}^c}}}{{y{\kern 1pt} {{\left[ {1 + {{\left( {\frac{y}{b}} \right)}^c}} \right]}^{k + 1}}}}}&{b,c,k,y > 0}
\end{array}\]

\[\begin{array}{*{20}{c}}
{{F_Y}(y;b,c,k) = 1 - \frac{1}{{{{\left[ {1 + {{\left( {\frac{y}{b}} \right)}^c}} \right]}^k}}}}&{b,c,k,y > 0}
\end{array}\]

We now consider the limiting distribution when $c \to \infty$ and $k \to 0$ in such a way that the product $ck$ remains constant, i.e. $ck = \lambda$.

Now, 
\[\begin{array}{l}
\mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \left\{ {{F_Y}(y;b,c,k)} \right\} = 1 - \mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \frac{1}{{{{\left[ {1 + {{\left( {\frac{y}{b}} \right)}^c}} \right]}^k}}}\\
\\
and\\
\\
\mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } {\left[ {1 + {{\left( {\frac{y}{b}} \right)}^c}} \right]^k} = \mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \left\{ {{{\left( {\frac{y}{b}} \right)}^{ck}}{{\left[ {1 + {{\left( {\frac{b}{y}} \right)}^c}} \right]}^k}} \right\}\\
\\
and\\
\\
\mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \left\{ {{{\left( {\frac{y}{b}} \right)}^{ck}}{{\left[ {1 + {{\left( {\frac{b}{y}} \right)}^c}} \right]}^k}} \right\} = \mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \left\{ {{{\left( {\frac{y}{b}} \right)}^{ck}}} \right\}\mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \left\{ {{{\left[ {1 + {{\left( {\frac{b}{y}} \right)}^c}} \right]}^k}} \right\}\\
 = \mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \left\{ {{{\left( {\frac{y}{b}} \right)}^{ck}}} \right\}\; \cdot \,1\\
 = {\left( {\frac{y}{b}} \right)^\lambda }
\end{array}\]

Therefore, \[\begin{array}{*{20}{c}}
{\mathop {\mathop {\lim }\limits_{(c,k) \to (\infty ,0)} }\limits_{ck = \lambda } \left\{ {{F_Y}(y;b,c,k)} \right\} = 1 - {{\left( {\frac{b}{y}} \right)}^\lambda }}&{y \ge b}
\end{array}\]

which we recognise as the (American) Pareto distribution. So, if the limiting distribution of $Y = \frac{1}{X}$ is a Pareto distribution, then the limiting distribution of $X = \frac{1}{Y}$ is the (American) *inverse Pareto* distribution:

\[\begin{array}{l}
{f_X}\left( {x;\alpha ,\beta } \right) = \lambda {b^\lambda }{x^{\lambda  - 1}};{\rm{  }}0 \le x \le {\textstyle{1 \over b}};{\rm{  }}\lambda {\rm{,}}b > 0\\
{F_X}\left( {x;\alpha ,\beta } \right) = {\left( {xb} \right)^\lambda };{\rm{  }}0 \le x \le {\textstyle{1 \over b}};{\rm{  }}\lambda {\rm{,}}b > 0
\end{array}\]

For completeness, the MLEs of this distribution have closed-form expressions and are given by:
\[\begin{array}{l}
\hat \lambda  = {\left[ {\ln \left( {\frac{{{g_X}}}{{\hat b}}} \right)} \right]^{ - 1}}\\
\hat b = \frac{1}{{\max \left\{ {{X_i}} \right\}}}{\rm{ }}
\end{array}\]

and ${\rm{ }}{g_X}$is the *geometric mean* of the data.

### Inverse Weibull distribution

<br>Let $X \sim Burr(b,c,k)$ have the *pdf* given in the box above. We make the transformation \[Y = \frac{{b{\kern 1pt} {k^{\tfrac{1}{c}}}{\kern 1pt} \theta }}{X}\]
where $\theta$ is a parameter (constant). The distribution of $Y$ is also a Burr distribution and has *cdf* \[{G_Y}\left( y \right) = 1 - \frac{1}{{{{\left[ {1 + {{\left( {\frac{y}{{{k^{\tfrac{1}{c}}}{\kern 1pt} \theta }}} \right)}^c}} \right]}^k}}}\].<br><br>We are interested in the limiting behaviour of this Burr distribution as $k \to \infty$.
<br>Now,\[\mathop {\lim }\limits_{k \to \infty } {G_Y}\left( y \right) = 1 - \mathop {\lim }\limits_{k \to \infty } {\left[ {1 + {{\left( {\frac{y}{{{k^{\tfrac{1}{c}}}{\kern 1pt} \theta }}} \right)}^c}} \right]^{ - k}}\]

\[{ = 1 - \mathop {\lim }\limits_{k \to \infty } {{\left[ {1 + \frac{{{{\left( {\frac{y}{\theta }} \right)}^c}}}{{k{\kern 1pt} }}} \right]}^{ - k}}}\]

\[\begin{matrix}
   =1-\exp \left[ -{{\left( \frac{y}{\theta } \right)}^{c}} \right]  \\
   \left\{ \text{using the fact that }\underset{n\to \infty }{\mathop{\lim }}\,{{\left( 1+{}^{z}\!\!\diagup\!\!{}_{n}\; \right)}^{-n}}={{e}^{-z}} \right\}  \\
\end{matrix}\]

We recognise the last expression as the *cdf* of a Weibull distribution with parameters $c$ and $\theta$.



```{r, results = "asis", echo = FALSE}
cat(ssdtools::ssd_licensing_md())
```
