---
title: "Customising Plots"
author: "ssdtools Team"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
bibliography: references.bib
# csl: style.csl
mainfont: Arial
mathfont: Courier
latex_engine: MathJax
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Customising Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Plotting the cumulative distributions

The `ssdtools` package plots the cumulative distribution functions using `ssd_plot_cdf()`. 

For example, consider the CCME boron data from the [`ssddata`](https://github.com/open-AIMS/ssddata) package. 
We can fit, and then plot the cdfs as follows.

```{r warning=FALSE, message=FALSE}
library(ssdtools)
library(ggplot2)

fits <- ssd_fit_dists(ssddata::ccme_boron)
ssd_plot_cdf(fits)
```

This graphic is a [ggplot](https://ggplot2.tidyverse.org) object and so can be customized in the usual way.

For example, we can add the model-averaged cdf by setting `average = NA` and then customize the color scale with `scale_color_manual()` and theme.

```{r}
ssd_plot_cdf(fits, average = NA) + 
  scale_color_manual(name = "Distribution", breaks = c("average", names(fits)), values = 1:7) +
  theme_bw()
```

## ggplot Geoms

The `ssdtools` package provides four ggplot geoms to allow you construct your own plots.

The first is `geom_ssdpoint()` which plots species sensitivity data

```{r}
ggplot(ssddata::ccme_boron) +
  geom_ssdpoint(aes(x = Conc)) +
  ylab("Probability density") +
  xlab("Concenration")
``` 

The second is `geom_ssdsegments()` which plots the range of censored species sensitivity data

```{r}
ggplot(ssddata::ccme_boron) +
  geom_ssdsegment(aes(x = Conc, xend = Conc * 2)) +
  ylab("Probability density") +
  xlab("Concenration")
``` 

The third is `geom_xribbon()` which plots species sensitivity confidence intervals

```{r}
ggplot(boron_pred) +
  geom_xribbon(aes(xmin = lcl, xmax = ucl, y = proportion)) +
  ylab("Probability density") +
  xlab("Concenration")
```  

And the fourth is `geom_hcintersect()` which plots hazard concentrations

```{r}
ggplot() +
  geom_hcintersect(xintercept = c(1, 2, 3), yintercept = c(0.05, 0.1, 0.2)) +
  ylab("Probability density") +
  xlab("Concenration")
```    

They can be combined together as follows

```{r}
gp <- ggplot(boron_pred, aes(x = est)) +
  geom_xribbon(aes(xmin = lcl, xmax = ucl, y = proportion), alpha = 0.2) +
  geom_line(aes(y = proportion)) +
  geom_ssdsegment(data = ssddata::ccme_boron, aes(x = Conc / 2, xend = Conc * 2)) +
  geom_ssdpoint(data = ssddata::ccme_boron, aes(x = Conc / 2)) +
  geom_ssdpoint(data = ssddata::ccme_boron, aes(x = Conc * 2)) +
  scale_y_continuous("Species Affected (%)", labels = scales::percent) +
  expand_limits(y = c(0, 1)) +
  xlab("Concentration (mg/L)")
print(gp + geom_hcintersect(xintercept = ssd_hc(fits)$est, yintercept = 5 / 100))
```

To log the x-axis add the following code.

```{r}
gp <- gp + coord_trans(x = "log10") +
  scale_x_continuous(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = comma_signif
  )
print(gp + geom_hcintersect(xintercept = ssd_hc(fits)$est, yintercept = 0.05))
```

The most recent plot can be saved as a file using `ggsave()`, which also allows the user to set the resolution.

```r
ggsave("file_name.png", dpi = 600)
```

## Fitting and plotting distributions to multiple groups such taxa and/or chemicals

An elegant approach using some tidyverse packages is demonstrated below.

```{r, message=FALSE}
library(ssddata)
library(ssdtools)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)

boron_preds <- nest(ssddata::ccme_boron, data = c(Chemical, Species, Conc, Units)) %>%
  mutate(
    Fit = map(data, ssd_fit_dists, dists = "lnorm"),
    Prediction = map(Fit, predict)
  ) %>%
  unnest(Prediction)
```

The resultant data and predictions can then be plotted as follows.

```{r}
library(ssdtools)
library(ssddata)
ssd_plot(ssddata::ccme_boron, boron_preds, xlab = "Concentration (mg/L)", ci = FALSE) +
  facet_wrap(~Group)
```


## Embellishing Plots with an Exposure Distribution

For example, suppose we want to superimpose an environmental concentration cumulative distribution and compute the exposure risk as outlined in @verdonck_limitations_2003.

```{r echo=FALSE}
# what are the mean and sd on the log-scale for the environmental concentration
ex.mean.log <- log(0.1)
ex.sd.log <- 1
```

Finding a suitable probability distribution to describe the exposure concentration is beyond the scope of this document -- we will assume that this has been done elsewhere.
In particular, suppose that the exposure concentration follows a log-normal distribution with a mean of `r signif(ex.mean.log, 3)` and a standard deviation of `r ex.sd.log` on the logarithmic scale. 
From the exposure distribution, we construct a data frame with the concentration values and the cumulative probability of seeing this exposure or less in the environment.

Notice that some care is needed because the ssdtools plot is on the logarithmic base 10 scale and not the natural logarithm base $e$ scale.

```{r}
ex.cdf <- data.frame(Conc = exp(seq(log(.01), log(10), .1))) # generate a grid of concentrations
ex.cdf$ex.cdf <- plnorm(ex.cdf$Conc,
  meanlog = ex.mean.log,
  sdlog = ex.sd.log
) # generate the cdf
```

We now add this to the plot
```{r}
gp +
  geom_line(data = ex.cdf, aes(x = Conc, y = ex.cdf), color = "red", linewidth = 2) +
  annotate("text",
    label = paste("Exposure distribution"),
    x = 1.08 * ex.cdf$Conc[which.max(ex.cdf$ex.cdf > 0.5)], y = 0.5, angle = 75
  )
```

The `ssdtools` package contains a function `ssd_exposure()` that computes the risk as defined by Verdonck et al (2003) representing the average proportion of species at risk.

```{r echo=TRUE}
set.seed(99)
ex.risk <- ssd_exposure(fits, meanlog = ex.mean.log, sdlog = ex.sd.log)
ex.risk
```

The risk of `r round(ex.risk, 5)` can also be added to the plot in the usual way:

```{r echo=TRUE}
gp +
  geom_line(dat = ex.cdf, aes(x = Conc, y = ex.cdf), color = "red", linewidth = 2) +
  annotate("text",
    label = paste("Exposure distribution"),
    x = 1.08 * ex.cdf$Conc[which.max(ex.cdf$ex.cdf > 0.5)], y = 0.5, angle = 75
  ) +
  annotate("text",
    label = paste("Verdonck risk :", round(ex.risk, 5)),
    x = Inf, y = 0, hjust = 1.1, vjust = -.5
  )
```

## References

<div id="refs"></div>

```{r, results = "asis", echo = FALSE}
cat(ssdtools::ssd_licensing_md())
```
