---
title: "Distributions"
author: "ssdtools Team"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 4,
  fig.width = 6
)
```

```{r setup, message=FALSE}
library(ssdtools)
```

## Default Distributions

By default, `ssdtools` fits three distributions to species sensitivity data:
the Burr Type-III 2-parameter, gamma and log-normal distributions.

### Default Hazard Concentrations

The hazard concentrations for the three default distributions with their default parameter values are plotted below.
```{r}
ssd_plot_cdf(list(burrIII2 = NULL, gamma = NULL, lnorm = NULL))
```

### Equally Distributed Hazard Concentrations

For comparative purposes we also plot the three default distributions with a mean of 0 and standard deviation of 1 on the logarithmic scale.
```{r}
ssd_plot_cdf(list(burrIII2 = NULL, gamma = NULL, lnorm = list(meanlog = 0, sdlog = 1)))
```

## All Distributions

For completeness the hazard concentrations for all the distributions that have been tested with `ssdtools` are plotted below with their default parameter values (by default the gamma and weibull hazard concentrations are identical).
```{r}
ssd_plot_cdf(list(burrIII2 = NULL, burrIII3 = NULL, gamma = NULL,
                  gompertz = NULL, lgumbel = NULL, llogis = NULL,
                  lnorm = NULL, pareto = NULL, weibull = NULL))

```