---
title: "Frequently Asked Questions"
author: "ssdtools Team"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  ```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

```{r}
library(ssdtools)
library(ggplot2)
```

## How do I fit a custom distribution?

The user can also specify one or more custom distributions (called say `custom`) provided vectorized density (`dcustom()`), distribution (`pcustom()`), quantile (`qcustom()`) and starting value (`scustom()`) functions are defined. 
For example see the [llogis source code](https://github.com/bcgov/ssdtools/blob/master/R/llogis.R).

## How do I fit distributions to multiple groups such taxa and/or chemicals?

An elegant approach using some tidyverse packages is demonstrated below.

```{r, message=FALSE}
library(purrr)
library(tidyr)
library(dplyr)

boron_preds <- nest(ssdtools::boron_data, data = c(Chemical, Species, Conc, Units)) %>%
  mutate(Fit = map(data, ssd_fit_dists, dists = "lnorm"),
         Prediction = map(Fit, predict)) %>% 
  unnest(Prediction)

boron_hc5s <- filter(boron_preds, percent == 5)
```

The resultant data and predictions can then be plotted as follows.
```{r, fig.width = 5, fig.height = 5}
ssd_plot(boron_data, boron_preds, xlab = "Concentration (mg/L)") +
  facet_wrap(~Group) +
  geom_hcintersect(data = boron_hc5s, aes(xintercept = est, yintercept = percent / 100))
```

### How do I plot a Cullen Frey plot?

The data can be visualized using a Cullen Frey plot of the skewness and kurtosis.

```{r, fig.width = 5, fig.height = 5}
ssd_plot_cf(boron_data)
```

## How do I plot model diagnostics?

A `fitdists` object can be plotted to display model diagnostics plots for each fit.
```{r, fig.width=6, fig.height=6, fig.show='hold'}
plot(boron_dists)
```
