---
title: "Frequently Asked Questions"
author: "ssdtools Team"
date: '`r format(Sys.time(), "%Y-%m-%d", tz = "UTC")`'
bibliography: references.bib
mathfont: Courier
latex_engine: MathJax
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

### How can I plot the model averaged fit and the individual fits?

```{r, fig.alt = "A plot showing the model averaged fit and the individual fits for the ccme_boron data."}
library(ssdtools)

dist <- ssd_fit_dists(ssddata::ccme_boron)
ssd_plot_cdf(dist, average = NA)
```

### How can I include mathematical expressions in the x-axis label when using `ssd_plot()`?

```{r, message = FALSE, fig.alt = "A plot showing the model averaged fit with a mathematical expression in the xlabel with the units."}
library(ssdtools)

ssd_plot(ssddata::ccme_boron, ssdtools::boron_pred, label = "Species", shape = "Group") +
  ggplot2::scale_x_continuous(latex2exp::TeX("Boron $(\\mu g$/L)$"), breaks = c(1, 10, 100, 1000, 10000))
```

<div id="refs"></div>

```{r, results = "asis", echo = FALSE}
cat(ssd_licensing_md())
```

## Censoring

### What is censored data?

Data is censored when the exact value is uncertain. 
There are three primary types of censoring.
Right censoring when only the lower bound is known (>), left censoring when only the upper bound is known (<)
and interval censored when both the lower and upper bound are known (><).
Uncensored data can be thought of as interval censored data in which the upper and lower bound are identical (=).
Likewise, missing data can be considered to be censored data in which both the upper and lower bound are unknown (?).

### How is censored data coded in ssdtools?

Censored data is coded by specifying the lower bound values in one column (>) and the upper bound values in a second column (<).
Unknown lower bounds are coded using NA or 0 while unknown upper bounds are coded using NA or Inf.
The name of the columns with the lower and upper bound values are then passed to the `left` (>) and `right` (<) arguments in the `ssd_fit_dists()` function, respectively.

### What are the current limitations on analysing censored data?

Currently `ssdtools` has the following limitations on the analysis of censored data:

- distribution fitting to left (<) or interval (><) censored data only, i.e., when the upper bound is known;
- model averaging when the distributions have the same number of parameters only; and
- estimation of confidence interval using non-parametric bootstrapping only.
