---
title: "Introduction to ssdca"
author: "Joe Thorley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Vignette Title}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 3,
  fig.height = 3
)
library(ssdca)
```


## Introduction

The `ssdca` package fits and plots Species Sensitivity Distributions.
This vignette describes how to use `ssdca` in conjunction with 
the `fitdistrplus` and `ggplot2` packages, which it installs and extends.

## Inputting Data

Data can be input from excel spreadsheets using `readxl::read_excel()` 
and from csv files using `readr::read_csv()`.

## Cullen Frey Plots

The data can be visualized using a cullen frey plot of the skewness and kurtosis.
Here we plot species sensitivity data for Boron.

```{r, fig.width = 5, fig.height = 5}
ssd_cfplot(boron_data)
```

## Fitting Distributions

The function `ssd_fit_dists()` inputs a data frame (with the concentration values, 
by default, in the `Conc` column) and fits one or more distributions.
Here we fit the log-normal and weibull distributions to the Boron data.
```{r}
library(ssdca)
boron_data
boron_dists2 <- ssd_fit_dists(boron_data, dists = c("lnorm", "weibull"))
boron_dists2
```

The coefficients can be extracted using the `coef` function
```{r}
lapply(boron_dists2, coef)
```

## Model Diagnostics and Goodness of Fit

The resultant `fitdists` object can be plotted to display model diagnostics plots for each fit.
```{r, fig.width=6, fig.height=6, fig.show='hold'}
plot(boron_dists2)
```

In addition, several goodness of fit statistics including the Akaike's Information Criteron corrected for sample size (`aicc`) can be generated using `ssd_gof()`.

Moving forwards we consider the same six fits to Boron as Schwarz and Tillmanns (2017).
```{r}
boron_dists <- ssd_fit_dists(boron_data)
ssd_gof(boron_dists)
```
For interpretation of the values see Schwarz and Tillmanns (2017).

## Plotting Fits

The actual fits can be plotted using the ggplot generic `autoplot`.
```{r, fig.width = 5 }
autoplot(boron_dists)
```

As `autoplot` returns a `ggplot` object it can be modified prior to plotting, 
which is extremely useful.
```{r, fig.width = 5 }
autoplot(boron_dists) + theme_bw()
```

## 5% Hazard Concentration

Typically, species sensitivity distributions are fitted to estimate the 5% hazard concentration ($HC_5$).
This is the concentration that affects just 5% of the species tested.
The model-averaged $HC_5$ estimate (with 95% confidence limits) can be calculated as follows
```{r}
ssd_hc5(boron_dists)
```
By default model averaging is based on the Akaike's Weights corrected for small sample size.
To return the individual $HC_5$ estimates and their respective Akaike's Weights set `average = FALSE`
```{r}
ssd_hc5(boron_dists, average = FALSE)
```
For interpretation of the values see Schwarz and Tillmanns (in prep.).

## Predictions

The `predict` function can be used to generate model-averaged (or individual) 
estimates across the full range of hazard concentrations.
```{r, eval = FALSE}
boron_pred <- predict(boron_dists)
```

The resultant values provide valuable information
```{r}
boron_pred
```
which can be plotted together with the original data to summarise an analysis
```{r, fig.width = 6, fig.height = 4}
theme_set(theme_bw()) # change the theme
ssd_plot(boron_data, boron_pred, shape = "Group", color = "Group", label = "Species",
         ylab = "Concentration (mg/L)") + 
  expand_limits(x = 5000) + # to ensure the species labels fit
  scale_color_manual(values = c("Amphibian" = "Black", "Fish" = "Blue", 
                                "Invertebrate" = "Red", "Plant" = "Brown")) +
  ggtitle("Species Sensitivity for Boron")
```

Once again the returned object is a `ggplot` object which can be customized prior to plotting.
By default, the plot includes the model-averaged 95% confidence interval as a shaded band and 
the model-averaged $HC_5$ as a dotted line.

## Weighted Data

The `ssd_fit_dists()` function allows the user to name a column that specifies
the weights to use in the estimation. 
However, to date very little functionality has been implemented for weighted fits in the `fitdistrplus` package.
```{r, error = TRUE}
boron_data$Weight <- as.integer(boron_data$Group)
fit <- ssd_fit_dists(boron_data, weight = "Weight", dists = c("lnorm", "weibull"))
fit
plot(fit)
```

## Censored Data

Censored data is that for which only a lower and/or upper limit is known for a particular species. 
If the `right` argument in `ssd_fit_dists()` is different to the `left` argument then the data are considered to be censored. 
`fluazinam` is a censored data set from the `fitdistrplus` package.

```{r}
data(fluazinam)
head(fluazinam)
```

There are less goodness-of-fit statistics available for
fits to censored data (currently just aic and bic).
```{r}
fluazinam_dists <- ssd_fit_dists(fluazinam)
ssd_gof(fluazinam_dists)
```
But model-averaged predictions can be made using AIC
```{r, eval = FALSE}
fluazinam_pred <- predict(fluazinam_dists)
```
and the results plotted complete with arrows indicating the censorship.
```{r, fig.width=5}
ssd_plot(fluazinam, fluazinam_pred, 
         left = "left", right = "right", 
         ylab = "Concentration (mg/L)")
```

## Extending ggplot

The `sscda` package provides three ggplot geoms to allow you construct your own plots.
They are `geom_ssd()` to plot the original data, 
`geom_hc5()` to plot the $HC_5$ and `geom_xribbon()` to plot the confidence intervals.

They can be used as follows

```{r, fig.width = 5, fig.height = 4}
gp <- ggplot(boron_pred, aes_string(x = "est")) + 
  geom_xribbon(aes_string(xmin = "lcl", xmax = "ucl", y = "prop"), alpha = 0.2) +
  geom_line(aes_string(y = "prop")) +
  geom_hc5(xintercept = boron_pred$est[boron_pred$prop == 0.05]) +
  geom_ssd(data = boron_data, aes_string(x = "Conc"))
print(gp)
```    

To log the x-axis use the following code.
```{r, fig.width = 5, fig.height = 4}
gp <- gp + coord_trans(x = "log10") +
  scale_x_continuous(breaks = scales::trans_breaks("log10", function(x) 10^x),
                     labels = comma_signif)
print(gp)
```

The current plot can be saved as a file using `ggsave()`,
which also allows the user to set the resolution.
```r
ggsave("file_name.png", dpi = 600)
```
