---
title: "Introduction to ssdca"
author: "Joe Thorley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Vignette Title}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ssdca)
```

The `ssdca` package fits Species Sensitivity Distributions and plots and tabulates the results.

The core function `ssd_fit_dist()` takes a numeric vector of sensitivity values
and fits a single distribution (by default the log-normal).
For a vector of all the recognized distributions type `sdd_dists()`.

```{r}
boron <- dplyr::filter(ccme_data, Chemical == "Boron")
dist <- ssd_fit_dist(boron$Conc)
print(dist)
```

The `ssd_fit_dist()` function returns a object of class `fitdist` as recognized by 
the `fitdistrplus` package. The `fitdistrplus` package is automatically installed 
(and loaded) when you install (and load) `ssdca`.
The base `plot()` generic function plots model diagnostics for `fitdist` objects.
```{r, fig.width=6, fig.height=6}
plot(dist)
```

The more familiar xx plots can be produced using ggplot2
```{r, fig.width=6, fig.height=3}
library(ggplot2)
library(dplyr)
library(scales)

ggplot(filter(ccme_data, Chemical %in% c("Boron", "Chloride")), aes(Conc)) +
  facet_wrap(~Chemical) +
  geom_ssd() +
  stat_fitdist(color = "red") +
  stat_fitdist(color = "blue", dist = "gamma") +
  coord_trans(x = "log10") +
  scale_x_continuous("Concentration", labels = comma, breaks = c(1,10,10^2,10^3,10^4,10^5)) +
  scale_y_continuous("Distribution (%)", labels = percent, limits = c(0 , 1), expand = c(0,0))
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

output: 
rmarkdown::html_vignette:
css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

output:
rmarkdown::html_vignette:
fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
