---
title: "Introduction to ssdca"
author: "Joe Thorley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Vignette Title}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ssdca)
```

The `ssdca` package fits Species Sensitivity Distributions and plots and tabulates the results.

The core function `ssd_fit_dists()` takes a numeric vector of sensitivity values
and fits one or more recognized distributions.

```{r}
boron <- ssd_fit_dists(ccme_data$Concentration[ccme_data$Chemical == "Boron"], c("lnorm", "weibull"))
print(boron)
```

There are currently `r length(ssd_dists(all = TRUE))` recognized distributions.
```{r}
ssd_dists(all = TRUE)
```

The base `plot()` generic function plots model diagnostics for `fitdists` objects.
```{r, fig.width=6, fig.height=6, fig.show='hold'}
plot(boron)
```

The more familiar xx plots can be produced using the ggplot2 generic `autoplot()`
```{r, fig.width=4, fig.height=3}
library(ggplot2)
autoplot(boron)
```

