---
title: "ssdtools User Manual"
author: "Joe Thorley"
date: "`r Sys.Date()`"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ssdtools User Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4
)
```

## Introduction

`ssdtools` is an R package to fit Species Sensitivity Distributions (SSDs) using Maximum Likelihood and model averaging.

SSDs are cumulative probability distributions that are used to estimate the percent of species that are affected by a given concentration of a chemical.
The concentration that affects 5% of the species is referred to as the 5% Hazard Concentration (HC).
For more information on SSDs the reader is referred to @posthuma_species_2001.

In order to use `ssdtools` you need to install R (see below) or use the Shiny [app](https://bcgov-env.shinyapps.io/ssdtools/).
The shiny app includes a user guide. 
This vignette is a user manual for the R package.

## Philosophy

`ssdtools` provides the key functionality required to fit SSDs using Maximum Likelihood and model averaging in R.
It is intended to be used in conjunction with [tidyverse](https://www.tidyverse.org) packages such as `readr` to input data, `tidyr` and `dplyr` to group and manipulate data and `ggplot2` [@ggplot2] to plot data.
As such it endeavours to fulfill the tidyverse [manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html).

## Installing

In order to install R [@r] the appropriate binary for the users operating system should be downloaded from [CRAN](https://cran.r-project.org) and then installed.

Once R is installed, the `ssdtools` package can be installed (together with the tidyverse) by executing the following code at the R console
```r
install.packages("ssdtools")
install.packages("tidyverse")
```

The `ssdtools` package (and key tidyverse packages) can then be loaded into the current session using
```{r}
library(ssdtools)
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(purrr)
```

## Getting Help

To get additional information on a particular function just type `?` followed by the name of the function at the R console.
For example `?ssd_gof` brings up the R documentation for the `ssdtools` goodness of fit function.

For more information on using R the reader is referred to [R for Data Science](https://r4ds.had.co.nz) [@wickham_r_2016].

If you discover a bug in `ssdtools` please file an issue with a [reprex](https://reprex.tidyverse.org/articles/reprex-dos-and-donts.html) (repeatable example) at <https://github.com/bcgov/ssdtools/issues>.

## Inputting Data

Once the `ssdtools` package has been loaded the next task is to input some data.
An easy way to do this is to save the concentration data for a *single* chemical as a column called `Conc` in a comma separated file (`.csv`). 
Each row should be the sensitivity concentration for a separate species.
If species and/or group information is available then this can be saved as `Species` and `Group` columns.
The `.csv` file can then be read into R using the following
```{r, eval = FALSE}
data <- read_csv(file = "path/to/file.csv")
```
For the purposes of this manual we use the CCME dataset for boron which is provided with the `ssdtools` package.
```{r}
data <- ssdtools::boron_data
print(data)
```

## Fitting Distributions

The function `ssd_fit_dists()` inputs a data frame and fits one or more distributions.
The user can specify a subset of the 

- [log-normal](https://en.wikipedia.org/wiki/Log-normal_distribution) (`lnorm`),
- [log-logistic](https://en.wikipedia.org/wiki/Log-logistic_distribution) (`llog`),
- [log-Gumbel](https://en.wikipedia.org/wiki/Gumbel_distribution) (`lgumbel`),
- [Gompertz](https://en.wikipedia.org/wiki/Gompertz_distribution) (`gompertz`),
- [gamma](https://en.wikipedia.org/wiki/Gamma_distribution) (`gamma`) and 
- [Weibull](https://en.wikipedia.org/wiki/Weibull_distribution) (`weibull`) 

distributions and/or include the 

- [Pareto](https://en.wikipedia.org/wiki/Pareto_distribution) (`pareto`)

distribution using the `dists` argument.
```{r}
dists <- ssd_fit_dists(data, dists = c("lnorm", "gompertz"))
```

The coefficients can be extracted using the `coef` function.
However, in and off themselves the coefficients are not that helpful.
```{r}
coef(dists)
```

It is generally much more informative to plot the fits using the `autoplot` generic function.
As `autoplot` returns a `ggplot` object it can be modified prior to plotting (printing) to make it look prettier.
```{r, fig.width = 5}
theme_set(theme_bw()) # set plot theme
gp <- autoplot(dists) 
gp <- gp + ggtitle("Species Sensitivity Distributions for Boron")
print(gp)
```

## Selecting One Distribution

Given multiple distributions the user is faced with choosing the best fitting distribution (or as discussed below averaging the results weighted by the fit).

For illustrative purposes we consider the same six distributions as @schwarz_improving_2019
```{r}
boron_dists <- ssd_fit_dists(boron_data)
gof <- ssd_gof(boron_dists)
gof[order(gof$delta),]
```

The `ssd_gof()` function returns several goodness of fit measures that can be used to select the best distribution including three statistics 

- [Anderson-Darling](https://en.wikipedia.org/wiki/Anderson–Darling_test) (`ad`) statistic,
- [Kolmogorov-Smirnov](https://en.wikipedia.org/wiki/Kolmogorov–Smirnov_test) (`ks`) statistic and
- [Cramer-von Mises](https://en.wikipedia.org/wiki/Cramér–von_Mises_criterion) (`cvm`) statistic

and three information criteria

- Akaike's Information Criterion (`aic`),
- Akaike's Information Criterion corrected for sample size (`aicc`) and
- Bayesian Information Criterion (`bic`)

Following @burnham_model_2002 we recommend the `aicc` for model selection.
The best fitting model is that with the lowest `aicc` (indicated by the model with a `delta` value of 0.000 in the goodness of fit table).
In the current example the best fitting model is the Gompertz distribution.

For further information on the advantages of an information theoretic approach in the context of selecting SSDs the reader is referred to @schwarz_improving_2019

## Averaging Multiple Distributions

Often other distributions will fit the data almost as well as the best distribution as evidenced by `delta` values < 2 [@burnham_model_2002]. 
In this situation the recommended approach is to estimate the average fit based on the relative weights of the distributions  [@burnham_model_2002].
The `aicc` based weights are indicated by the `weight` column in the goodness of fit table.
In the current example, the gompertz, gamma, Weibull and log-normal distributions have `delta` values < 2.

## Estimating the Fit

The `predict` function can be used to generate estimates model-averaged (or if `average = FALSE` individual) estimates.
By default model averaging is based on `aicc`.

```{r, eval = FALSE}
boron_pred <- predict(boron_dists)
```

The resultant object is a data frame of the estimated concentration (`est`) with standard error (`se`) and lower (`lcl`) and upper (`ucl`) 95% confidence limits by percent of species affected (`percent`).
The uncertainty in the estimates is generated using parametric bootstrapping.
```{r}
boron_pred
```
The data frame of the estimates can then be plotted together with the original data using the `ssd_plot()` function to summarize an analysis.
Once again the returned object is a `ggplot` object which can be customized prior to plotting.
```{r, fig.height = 5, fig.width = 6}
gp <- ssd_plot(boron_data, boron_pred, color = "Group", label = "Species",
         xlab = "Concentration (mg/L)", ribbon = TRUE)
gp <- gp + expand_limits(x = 5000) + # to ensure the species labels fit
  scale_color_manual(values = c("Amphibian" = "Black", "Fish" = "Blue", 
                                "Invertebrate" = "Red", "Plant" = "Brown")) +
  ggtitle("Species Sensitivity for Boron")
print(gp)
```

In the above plot the model-averaged 95% confidence interval is indicated by the shaded band and the model-averaged 5% Hazard Concentration ($HC_5$) by the dotted line.
Hazard concentrations are discussed below.

## Hazard Concentrations

The 5% hazard concentration ($HC_5$) is the concentration that affects 5% of the species tested.
It can be obtained by selecting the estimated prediction with a percent value of 5.
```{r}
boron_pred[boron_pred$percent == 5,]
```

By default the uncertainty in the predicted estimates is generated from 1,000 bootstrap iterations.
However in the case of a specific hazard concentration we recommend the use of 10,000 bootstrap iterations to ensure repeatability.
Rather than regenerate all the predicted estimates with 10,000 iterations which may be prohibitively time-consuming we recommend the use of `ssd_hc()` to generate the single estimate of interest with 10,000 iterations.

```{r, eval = FALSE}
boron_hc5 <- ssd_hc(boron_dists, nboot = 10000)
```

The code may still take upwards of several minutes to run.

```{r}
print(boron_hc5)
```

## Plotting

The `ssdtools` package provides three ggplot geoms to allow you construct your own plots.

The first is `geom_ssd()` which plots species sensitivity data 
```{r}
ggplot(boron_data) + 
  geom_ssd(aes_string(x = "Conc"))
``` 

The second is `geom_xribbon()` which plots species sensitivity confidence intervals
```{r}
ggplot(boron_pred) + 
  geom_xribbon(aes_string(xmin = "lcl", xmax = "ucl", y = "percent/100"))
```  

And the third is `geom_hcintersect()` which plots hazard concentrations
```{r}
ggplot() + 
  geom_hcintersect(xintercept = c(1,2,3), yintercept = c(5,10,20)/100) 
```    

They can be combined together as follows
```{r}
gp <- ggplot(boron_pred, aes_string(x = "est")) + 
  geom_xribbon(aes_string(xmin = "lcl", xmax = "ucl", y = "percent/100"), alpha = 0.2) +
  geom_line(aes_string(y = "percent/100")) +
  geom_ssd(data = boron_data, aes_string(x = "Conc")) +
  scale_y_continuous("Species Affected (%)", labels = scales::percent) +
  expand_limits(y = c(0, 1)) +
  xlab("Concentration (mg/L)")
print(gp + geom_hcintersect(xintercept = boron_hc5$est, yintercept = 5/100))
```

To log the x-axis add the following code.
```{r}
gp <- gp + coord_trans(x = "log10") +
  scale_x_continuous(breaks = scales::trans_breaks("log10", function(x) 10^x),
                     labels = comma_signif)
print(gp + geom_hcintersect(xintercept = boron_hc5$est, yintercept = 5/100))
```

The most recent plot can be saved as a file using `ggsave()`, which also allows the user to set the resolution.
```r
ggsave("file_name.png", dpi = 600)
```

## Grouping

A common question is how do I fit distributions to multiple groups such taxa and/or chemicals?
An elegant approach using the tidyverse is demonstrated below.

```{r}
boron_datas <- nest(boron_data, -Group)
boron_datas <- mutate(boron_datas, 
                      Fit = map(data, ssd_fit_dists, dists = "lnorm"), 
                      Prediction = map(Fit, predict))
boron_datas <- unnest(boron_datas, Prediction)
```

The resultant data and predictions can then be plotted as follows.
```{r, fig.width = 5, fig.height = 5}
boron_hc5s <- filter(boron_datas, percent == 5)
gp %+% boron_datas + 
  facet_wrap(~Group) + 
  geom_hcintersect(data = boron_hc5s, aes(xintercept = est, yintercept = percent/100))
```

## Additional Features

### Cullen Frey Plots

The data can be visualized using a cullen frey plot of the skewness and kurtosis.

```{r, fig.width = 5, fig.height = 5}
library(ssdtools)
ssd_cfplot(boron_data)
```

### Model Diagnostics

A `fitdists` [@fitdistrplus] object can be plotted to display model diagnostics plots for each fit.
```{r, fig.width=6, fig.height=6, fig.show='hold'}
plot(dists)
```

### Censored Data

Censored data is that for which only a lower and/or upper limit is known for a particular species. 
If the `right` argument in `ssd_fit_dists()` is different to the `left` argument then the data are considered to be censored. 
`fluazinam` is a censored data set from the `fitdistrplus` package.

```{r}
data(fluazinam, package = "fitdistrplus")
head(fluazinam)
```

There are less goodness-of-fit statistics available for
fits to censored data (currently just `aic` and `bic`).
The `delta` values are calculated using `aic`.

```{r}
fluazinam_dists <- ssd_fit_dists(fluazinam, left = "left", right = "right")
ssd_gof(fluazinam_dists)
```
But model-averaged predictions (and hazard concentrations complete with 95% confidence limits) can be calculated using `aic`
```{r, eval = FALSE}
fluazinam_pred <- predict(fluazinam_dists)
```
and the results plotted complete with arrows indicating the censorship.
```{r}
ssd_plot(fluazinam, fluazinam_pred, 
         left = "left", right = "right", 
         xlab = "Concentration (mg/L)")
```

## References
