---
title: "ssdtools User Manual"
author: "Joe Thorley"
date: "`r Sys.Date()`"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ssdtools User Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

`ssdtools` is an R package to fit Species Sensitivity Distributions (SSDs) using Maximum Likelihood and model averaging.

SSDs are cumulative probability distributions that are used to estimate the
proportion of a range of species that are affected by a given concentration of a chemical.
The concentration that affects 5% of the species is referred to as the 5% Hazard Concentration (HC).
For more information on SSDs the reader is referred to @posthuma_species_2001.

In order to use `ssdtools` you need to install R (see below) or use the Shiny [app](https://poissonconsulting.shinyapps.io/ssdtools/).
The shiny app includes a user guide. 
This vignette is a user manual for the R package.

## Installation

In order to install R [@r] the appropriate binary for the users operating system should be downloaded from [CRAN](https://cran.r-project.org) and then installed.

Once R is installed, the `ssdtools` package can be installed by executing the following code at the R console
```r
install.packages("ssdtools")
```
The `ssdtools` package can then be loaded into the current session using
```{r}
library(ssdtools)
```

The `ssdtool` package loads several other R packages including `ggplot2` [@ggplot2] for plotting 
and `fitdistrplus` [@fitdistrplus] for model fitting.

## Getting Help

To get additional information on a particular function just type `?` followed by the name of the function at the R console.
For example `?ssd_gof` brings up the R documentation for the `ssdtools` goodness of fit function.

## Input Data

Once the `ssdtools` package has been loaded the next task is to input some data.
An easy way to do this is to save the concentration data for a *single* chemical as a column called `Conc` in a comma separated file (`.csv`). 
Each row should be the sensitivity concentration for a separate species.
If species and/or group information is available then this can be saved as `Species` and `Group` columns.
The `.csv` file can then be read into R using 
```r
data <- read.csv(file = "path/to/file.csv")
```
For the purposes of this manual we use the CCME dataset for boron which is provided with the ssdtools package
```{r}
data <- ssdtools::boron_data
print(data)
```

## Fit Distributions

With the data loaded the user can use the `ssd_fit_dists()` function to fit one or more distributions.
By default `ssd_fit_dists()` fits the log-normal (`lnorm`), log-logistic (`llog`), log-Gumbel (`lgumbel`), Gompertz (`gompertz`), gamma (`gamma`) and Weibull (`weibull`) distributions.
The user can change specify a subset using the `dists` argument.
```{r}
dists <- ssd_fit_dists(data, dists = c("lnorm", "gompertz", "gamma", "weibull"))
```

The fits can be plotted using `autoplot`.
```{r, fig.width = 6, fig.height= 6}
library(ggplot2)
autoplot(dists)
```


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

## References

