---
title: "Small sample bias in estimates"
author: "sstools Team"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Small sample bias in estimates}   
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 6,
                      fig.height = 6)

library(ggplot2)
library(mle.tools)
library(reshape2)
library(ssddata)
library(ssdtools)

nsim <- 1000 # number of simulations for effect of bias correction
```

## Introduction

### What is the issue

The *ssdtools* package uses the method of Maximum Likelihood (ML) to estimate parameters for each distribution that is fit to the data. Statistical theory
says that maximum likelihood estimators are asymptotically unbiased, but does
not guarantee performance in small samples.

For example, consider the CCME silver data that ships with *ssdtools*.

```{r warning=FALSE, message=FALSE}
Ag <- ssddata::ccme_data[ssddata::ccme_data$Chemical=="Silver",]
Ag$ecdf <- (rank(Ag$Conc)+.25)/(nrow(Ag)+.5)
Ag
```

Let us fit a log-normal distribution to the Ag endpoint data and estimate the parameters:

```{r warning=FALSE, message=FALSE}
fit <- ssd_fit_dists(Ag, dist="lnorm")
fit
```

For most distributions, the MLE must be found numerically by iterative
methods, but the log-normal distribution has easily computed estimators.

The *meanlog* parameter shown above represents the mean of the concentrations
on the (natural) logarithmic scale and we can easily reproduce
this value:

```{r }
mean(log(Ag$Conc))
```

The *sdlog* parameter represents the standard deviation on the logarithmic scale, but
the direct computation of the standard deviation gives a slightly different
result:

```{r }
sd(log(Ag$Conc))
```


It turns out that in small samples, the MLE of the standard deviation
for a log-normal distribution has a negative bias, i.e. the MLE tends to be
smaller than the underlying true parameter value. The cause of this bias
is found by comparing the formula for the MLE of the standard deviation
and the traditional estimator for the standard deviation:

$$ \widehat{\sigma}_{MLE} = \sqrt{\frac{\sum{(Y_i-\overline{Y})^2}}{{n}}}$$
$$ \widehat{\sigma}_{traditional} = \sqrt{\frac{\sum{(Y_i-\overline{Y})^2}}{{n-1}}}$$

where $n$ is the sample size, $Y_i$ are the $\log(concentrations)$, and
$\overline{Y}$ is the sample mean concentration again on the logarithmic scale.

We notice that the MLE uses a divisor of $n$ while the traditional
method uses a divisor of $n-1$. Hence the MLE has a negative bias and its value 
is
`r round(fit$lnorm$estimate["sdlog"]/sd(log(Ag$Conc)),2)`x the 
usual estimator for $\sigma$ which is 
$\sqrt{\frac{n-1}{n}}$ = 
`r round(sqrt((nrow(Ag)-1)/(nrow(Ag))),2)` evaluated at $n$ = `r nrow(Ag)` . 

As the sample size
increases the absolute size of the bias will get smaller and smaller, i.e.,
if $n=20$, then the MLE estimator is `r round(sqrt((20-1)/(20)),2)`x the traditional 
estimator for $\sigma$
which is neglible given the uncertainties in the actual end points.

Conversly, as the sample size decreases, the absolute size of the bias could become quite large, i.e.,
if $n=4$, then the MLE is `r round(sqrt((4-1)/(4)),2)`x the traditional estimator. But
if you are fitting a species sensitivity distribution to only 4 data values, 
perhaps concern about bias in MLE is misplaced. Australian guidelines recommend a minimum sample size of 8 species.

## What is the impact on the HCx value?
If the standard deviation is underestimated, then the tails of the 
distribution will be pulled inwards and the HCx values will tend 
to be larger compared to the case where the standard deviation is not deflated 
as shown in the following plot:

```{r echo=FALSE,message=FALSE, warning=FALSE}
MLEcurve <- data.frame(logConc=seq(-3,3,.1), source="MLE")
MLEcurve$density <- pnorm(MLEcurve$logConc, mean=fit$lnorm$estimate["meanlog"],
                                    sd=  fit$lnorm$estimate["sdlog"])
REGcurve <- data.frame(logConc=seq(-3,3,.1), source="Corrected SD")
REGcurve$density <- pnorm(REGcurve$logConc,
                          mean=fit$lnorm$estimate["meanlog"],
                            sd=  sd(log(Ag$Conc)))
plotdata <- rbind(MLEcurve, REGcurve)
ggplot(data=plotdata, aes(x=logConc, y=density))+
   ggtitle( "Comparing the estimated cumulative density computed using \nMLE and bias-corrected SD",
            subtitle=paste("Ag CCME data with n=",nrow(Ag),sep=""))+
   geom_line(aes(color=source, linetype=source))+
   geom_hline(yintercept=.05)+
   ylab("Cumulative probabiity")+
   xlab("log(Concentration) \n Horizontal line represents the HC5")+
   geom_point(data=Ag, aes(x=log(Conc),y=ecdf))
   

```

```{r echo=FALSE}
# compute the HC5 from the two fit
MLE.HC5.log <- qnorm(.05,mean=fit$lnorm$estimate["meanlog"], sd=  fit$lnorm$estimate["sdlog"])
MLE.HC5     <- exp(MLE.HC5.log)
REG.HC5.log <- qnorm(.05,mean=fit$lnorm$estimate["meanlog"], sd= sd(log(Ag$Conc)) )
REG.HC5     <- exp(REG.HC5.log)

MLE.HC1.log <- qnorm(.01,mean=fit$lnorm$estimate["meanlog"], sd=  fit$lnorm$estimate["sdlog"])
MLE.HC1     <- exp(MLE.HC1.log)
REG.HC1.log <- qnorm(.01,mean=fit$lnorm$estimate["meanlog"], sd= sd(log(Ag$Conc)) )
REG.HC1     <- exp(REG.HC1.log)

    
```

The HC5 estimated from the MLE fit is `r round(MLE.HC5.log,2)` on the logarithmic concentrations
scale or `r round(MLE.HC5,3)` on the concentration scale.
The HC5 estimated after correcting the standard deviation for small sample bias is
`r round(REG.HC5.log,2)` on the logarithmic concentrations
scale or `r round(REG.HC5,3)` on the concentration scale. 
The ratio of HC5 values 
is `r round(REG.HC5/MLE.HC5,2)`x on the concentration scale, i.e.
the estimated HC5 from the MLE is `r round(MLE.HC5/REG.HC5,1)`x larger than the HC5 computed using the bias correction.

The differences between the HCx computed from the MLE fit and using the corrected
standard deviation will become more pronounced for small HCx values. For example,
the HC1 estimated from the MLE fit is `r round(MLE.HC1,3)` 
and the HC1 estimated using the corrected standard deviation
is `r round(REG.HC1,3)` on the concentration scale. 
The ratio of the HC1 values
is `r round(REG.HC1/MLE.HC1,2)`x on the concentrations scale, i.e.
the estimated HC1 from the MLE is `r round(MLE.HC1/REG.HC1,1)`x larger than the HC5 computed using the bias correction.


## What can be done?
A similar concernalso occurs with other distributions. Howeve, except for a few distributions,
such as the normal distribution, analytical expressions for the MLE
and for unbiased estimators do not exist. The *mle.tools* package from CRAN
provides a method that numerically corrects the bias after the fit is completed.

### Bias correction using Cox-Snell method - log-normal distribution

For example, again using the Ag log-normal fit we have:
```{r }
# apply the Cox and Snell (1968) bias correction using mle.tools.
# what is the density function
norm.pdf <- quote(1 / (sqrt(2 * pi) * sigma) * exp(-0.5 / sigma ^ 2 * (x - mu) ^ 2))
norm.pdf

# what is the log(density) function (ignoring constants)
log.norm.pdf <- quote(- log(sigma) - 0.5 / sigma ^ 2 * (x - mu) ^ 2)
log.norm.pdf


bias.correct <- coxsnell.bc(density = norm.pdf, 
            logdensity = log.norm.pdf, 
            n = length(Ag$Conc), 
            parms = c("mu", "sigma"),
            mle = c(fit$lnorm$estimate["meanlog"], 
                    fit$lnorm$estimate["sdlog" ]), 
            lower = '-Inf', upper = 'Inf')
bias.correct

```

The biased corrected value for the standard deviation is
`r round(bias.correct$mle.b["sigma"],2)` which is comparable 
to the standard deviation of the log(concentration) found
earlier of `r round(sd(log(Ag$Conc)),2)`.


A small simulation study was conducted to investigate the effect of sample size
on the bias correction and effects of the small-sample bias in the estimates
of HC5 and HC1. For this simulation study, it was assumed that a log-normal
distribution represented the distribution of endpoints among species with a mean of 0
and a standard deviation of 1 (on the logarithmic scale). These values are arbitrary, but
any log-normal distribution can be rescaled (e.g. by changing units) to have this mean 
and standard deviation. 

Simulated data sets at various sample sizes were generated, the MLE and bias-corrected estimates
were obtained and these were used to estimate the HC5 and HC1 on the $\log()$ and anti-log scales. The average
value of each response was then computed and plotted vs. the actual parameter values based
on the known mean and standard deviation (shown in the plot below as a black horizontal line). 
For example, for a log-normal distribution with a mean
of 0 and a standard deviation of 1 on the log-scale, the $log(HC5)$ is the 0.05 quantile
of the normal distribution or `r round(qnorm(.05, 0, 1),3)`. 

A plot of the results is:

```{r message=FALSE, warning=FALSE, include=FALSE}
# Simulation study to estimate effect of small sample bias on estimates of HC5 and HC1
# Use a normal distribution 
set.seed(964544)

mu <- 0
sd <- 1

sample.sizes <- c(6, 8, 10, 12, 15, 20, 25, 30, 50)

res <- plyr::ldply(sample.sizes, function(n, mu, sd, nsim=1000){
  fits <- plyr::ldply(1:nsim, function(sim, n, mu, sd){
     # generate sample of size n from a normal distribution with specified mean and variance
     #browser()
     df <- data.frame(Conc= exp(rnorm(n, mean=mu, sd=sd)))
     
     # find the mle using ssd tools 
     fit <- ssd_fit_dists(df, dist="lnorm")
     
     # find the bias corrected estimates
     norm.pdf <- quote(1 / (sqrt(2 * pi) * sigma) * exp(-0.5 / sigma ^ 2 * (x - mu) ^ 2))
     # what is the log(density) function (ignoring constants)
     log.norm.pdf <- quote(- log(sigma) - 0.5 / sigma ^ 2 * (x - mu) ^ 2)
     log.norm.pdf

     bias.correct <- coxsnell.bc(density = norm.pdf, 
            logdensity = log.norm.pdf, 
            n = n, 
            parms = c("mu", "sigma"),
            mle = c(fit$lnorm$estimate["meanlog"], 
                    fit$lnorm$estimate["sdlog" ]), 
            lower = '-Inf', upper = 'Inf')

     # find the hc5 and hc1 using the MLE and bias corrected values
     # compute the HC5 from the two fit
     actual.HC5.log <- qnorm(.05, mean=mu, sd=sd)
     actual.HC1.log <- qnorm(.01, mean=mu, sd=sd)
     actual.HC5     <- exp(actual.HC5.log)
     actual.HC1     <- exp(actual.HC1.log)
     MLE.HC5.log <- qnorm(.05,mean=fit$lnorm$estimate["meanlog"], sd=  fit$lnorm$estimate["sdlog"])
     MLE.HC5     <- exp(MLE.HC5.log)
     BC.HC5.log  <- qnorm(.05,mean=bias.correct$mle.bc["mu"], 
                             sd   =bias.correct$mle.bc["sigma"] )
     BC.HC5      <- exp(BC.HC5.log)

     MLE.HC1.log <- qnorm(.01,mean=fit$lnorm$estimate["meanlog"], sd=  fit$lnorm$estimate["sdlog"])
     MLE.HC1     <- exp(MLE.HC1.log)
     BC.HC1.log  <- qnorm(.01,mean=bias.correct$mle.bc["mu"], 
                             sd   =bias.correct$mle.bc["sigma"] )
     BC.HC1      <- exp(BC.HC1.log)

     data.frame(n=n, sim=sim,  mu=mu, sd=sd,
              mle.meanlog=fit$lnorm$estimate["meanlog"], 
              mle.sdlog  =fit$lnorm$estimate["sdlog"],
              bc.meanlog = bias.correct$mle.bc["mu"],
              bc.sdlog   = bias.correct$mle.bc["sigma"],
              actual.HC5.log,
              actual.HC5,
              actual.HC1.log,
              actual.HC1,
              MLE.HC5.log = MLE.HC5.log,
              MLE.HC5,
              BC.HC5.log,
              BC.HC5,
              MLE.HC1.log,
              MLE.HC1,
              BC.HC1.log,
              BC.HC1)
   }, n=n, mu=mu, sd=sd) 
   fits
}, mu=mu, sd=sd, nsim=nsim)

head(res)
# summarize the output from the simulation
res.summary <- plyr::ddply(res,"n", plyr::summarize,
                           n=mean(n),
                           nsims=max(sim),
                           mu=mean(mu),
                           sd=mean(sd),
                           mean.mle.meanlog= mean(mle.meanlog),
                           mean.mle.sdlog  = mean(mle.sdlog),
                           mean.bc.meanlog = mean(bc.meanlog),
                           mean.bc.sdlog   = mean(bc.sdlog),
                           actual.HC5.log  = mean(actual.HC5.log),
                           actual.HC1.log  = mean(actual.HC1.log),
                           mean.mle.HC5.log= mean(MLE.HC5.log),
                           mean.mle.HC1.log= mean(MLE.HC1.log),
                           mean.bc.HC5.log = mean(BC.HC5.log),
                           mean.bc.HC1.log = mean(BC.HC1.log),
                           mean.mle.HC5    = mean(MLE.HC5),
                           mean.mle.HC1    = mean(MLE.HC1),
                           mean.bc.HC5     = mean(BC.HC5),
                           mean.bc.HC1     = mean(BC.HC1)

                           )
res.summary


plotdata <- reshape2::melt(res.summary,
                           id.vars="n",
                           value.name="value",
                           variable.name="Measure")
plotdata$Measure <- as.character(plotdata$Measure)
unique(plotdata$Measure)
str(plotdata)

plotdata$parameter <- car::recode(plotdata$Measure,
                          " 'mean.mle.meanlog'='Mean of log(Conc)';
                            'mean.bc.meanlog' ='Mean of log(Conc)';
                            'mean.mle.sdlog'  ='SD of log(Conc)';
                            'mean.bc.sdlog'   ='SD of log(Conc)';
                            'mean.mle.HC5.log'='log(HC)';
                            'mean.mle.HC1.log'='log(HC)';
                            'mean.bc.HC5.log' ='log(HC)';
                            'mean.bc.HC1.log' ='log(HC)';
                            'mean.mle.HC5'    ='actual HC';
                            'mean.mle.HC1'    ='actual HC';
                            'mean.bc.HC5'     ='actual HC';
                            'mean.bc.HC1'     ='actual HC';

                          ")

plotdata$method <- car::recode(plotdata$Measure,
                           "'mean.mle.meanlog'='MLE';
                            'mean.bc.meanlog' ='BC';
                            'mean.mle.sdlog'  ='MLE';
                            'mean.bc.sdlog'   ='BC';
                            'mean.mle.HC5.log'='MLE';
                            'mean.mle.HC1.log'='MLE';
                            'mean.bc.HC5.log' ='BC';
                            'mean.bc.HC1.log' ='BC';
                            'mean.mle.HC5'    ='MLE';
                            'mean.mle.HC1'    ='MLE';
                            'mean.bc.HC5'     ='BC';
                            'mean.bc.HC1'     ='BC';

                          ")

plotdata$HC <- car::recode(plotdata$Measure,
                           "'mean.mle.HC5.log'='HC5';
                            'mean.mle.HC1.log'='HC1';
                            'mean.bc.HC5.log' ='HC5';
                            'mean.bc.HC1.log' ='HC1';
                            'mean.mle.HC5'    ='HC5';
                            'mean.mle.HC1'    ='HC1';
                            'mean.bc.HC5'     ='HC5';
                            'mean.bc.HC1'     ='HC1';

                           else='NA';
                          ")

head(plotdata)

xtabs(~Measure+parameter,  data=plotdata, exclude=NULL, na.action=na.pass)
xtabs(~Measure+method   ,  data=plotdata, exclude=NULL, na.action=na.pass)

select <- grepl("meanlog", plotdata$Measure) |
          grepl("sdlog"  , plotdata$Measure) |
          grepl("mle.HC5.log", plotdata$Measure) |
          grepl("mle.HC1.log", plotdata$Measure) |
          grepl("bc.HC5.log", plotdata$Measure) |
          grepl("bc.HC1.log", plotdata$Measure) |
          grepl("mle.HC5",    plotdata$Measure) |
          grepl("mle.HC1",    plotdata$Measure) |
          grepl("bc.HC5",     plotdata$Measure) |
          grepl("bc.HC1",     plotdata$Measure)

plotdata[select,]

true.parms <- data.frame(parameter=c("Mean of log(Conc)",
                                    "SD of log(Conc)",
                                    "log(HC)",
                                    "log(HC)",
                                    "actual HC",
                                    'actual HC'),
                         value=c(0,1,qnorm(.05, mu, sd), qnorm(.01, mu, sd), exp(qnorm(.05, mu, sd)), exp(qnorm(.01, mu, sd))),
                         linetype=c("solid","solid","dashed","solid","dashed","solid"), stringsAsFactors=FALSE)
true.parms

simplot <- ggplot(data=plotdata[select,], aes(x=n, y=value,color=method, linetype=HC))+
  ggtitle("Performance of mle and bias corrected estimators",
          subtitle=paste("Log-normal distribution with mean= ", mu, ' and sd =',sd,' on the log() scale',sep=""))+
  geom_point(position=position_jitter(w=0.4))+
  geom_line( position=position_jitter(w=0.4))+
  facet_wrap(~parameter, ncol=2,scales="free")+
  geom_hline(data=true.parms, aes(yintercept=value), linetype=true.parms$linetype)+
  xlab("Sample size")+
  scale_linetype_discrete(na.value="solid", breaks=c("HC1","HC5"), )
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
simplot
```

The MLE is unbiased for the mean of the log-normal distribution (bottom left plot)  -
the apparent deviations from the true value of 0 are very small (note the scale
on the $Y$ axis) and simply simulation artefacts. 

The MLE for the standard deviation
is biased downwards (lower right plot) and the bias become smaller with 
increasing sample size (the curve for the mean of the MLE estimate of the
standard deviation increases and approaches the true value of 0). The bias-correction for the
standard deviation is effective for all but the smallest sample sizes.

The estimated $\log{HC}$ (upper right plot) based on the MLE is biased upwards (i.e. larger) than the true
values but the bias declines with sample size (as expected). The 
estimate of the $\log{HC}$ based on the bias-corrected estimates performs well 
(close to the true value) except at very small sample sizes.

Finally, the estimated $HC1$ and $HC5$ values are again biased upwards (upper left plot). 
This bias consists of two parts  

1. bias in the underlying estimates of the parameters
of the distribution
2. non-linear tranformation bias, i.e. the mean of a function of the parameter values is
not equal to the function evaluated at the mean of the parameter values. For example, the
HC5 is found as the anti-log of the 5$^{th}$ percentile of the normal distribution. Suppose 
we have two simulation results where the estimated 5$^{th}$ percentile of the fitted normal distribution
were $-1.8$ and $-1.5$. The mean of the estimated 5$^th$ percentile is $\frac{-1.8+(-1.5)}{2}=-1.65$ and
is unbiased for the actual percentile value of $-1.645$. However, the actual HC5 is found as the anti-log
of the two individual estimates, i.e. $\exp(-1.8)=0.165$ and $\exp(-1.6)=.223$ whose mean is 0.194, but 
the anti-log of the average, $exp(-1.65)=.192$ which is not the same value.

The total bias does not appear to be large except in the case of very small sample sizes.

### Bias correction using Cox-Snell method - gamma distribution

We can also apply this to other distributions such as the gamma
distribution. If we fit a gamma distribution to the
Ag data we obtain:
```{r warnings=FALSE, message=FALSE}
fit.gamma <- ssd_fit_dists(Ag, dist="gamma")
fit.gamma
```
The bias corrected estimates are:
```{r }
# apply the Cox and Snell (1968) bias correction using mle.tools.
# what is the density function
gamma.pdf <- quote(1 /(scale ^ shape * gamma(shape)) * x ^ (shape - 1) * exp(-x / scale))
gamma.pdf

# what is the log(density) functiong ingoring constants
log.gamma.pdf <- quote(-shape * log(scale) - lgamma(shape) + shape * log(x) -
 x / scale)
log.gamma.pdf

bias.correct.gamma <- coxsnell.bc(density = gamma.pdf, 
            logdensity = log.gamma.pdf, 
            n = length(Ag$Conc), 
            parms = c("shape", "scale"),
            mle = c(fit.gamma$gamma$estimate["shape"], 
                    fit.gamma$gamma$estimate["scale" ]), 
            lower = 0, upper = 'Inf')
bias.correct.gamma

```

The two cumulative density functions are:

```{r echo=FALSE,message=FALSE, warning=FALSE}
MLEcurve <- data.frame(logConc=seq(-5,3,.1), source="MLE")
MLEcurve$density <- pgamma(exp(MLEcurve$logConc),  
                           shape=fit.gamma$gamma$estimate["shape"],
                           scale=fit.gamma$gamma$estimate["scale"])
REGcurve <- data.frame(logConc=seq(-5,3,.1), source="Bias corrected")
REGcurve$density <- pgamma(exp(REGcurve$logConc),
                           shape=bias.correct.gamma$mle.bc["shape"],
                           scale=bias.correct.gamma$mle.bc["scale"])
plotdata <- rbind(MLEcurve, REGcurve)
ggplot(data=plotdata, aes(x=logConc, y=density))+
   ggtitle( "Comparing the estimated cumulative gamma density \nusing MLE and bias correction",
            subtitle=paste("Ag CCME data with n=",nrow(Ag),
                            ' and gamma fit',sep=""))+
   geom_line(aes(color=source, linetype=source))+
   geom_hline(yintercept=.05)+
   xlab("log(Concentration) \n Horizontal line represents the HC5")+
   ylab("Cumulative probability")+
   geom_point(data=Ag, aes(x=log(Conc),y=ecdf))

```

```{r echo=FALSE}
# compute the HC5 from the two fit
MLE.gamma.HC5.log <- log(ssd_qgamma(.05, 
                           shape=fit.gamma$gamma$estimate["shape"],
                           scale=fit.gamma$gamma$estimate["scale"]))
MLE.gamma.HC5     <- exp(MLE.gamma.HC5.log)
REG.gamma.HC5.log <- log(ssd_qgamma(.05,
                           shape=bias.correct.gamma$mle.bc["shape"],
                           scale=bias.correct.gamma$mle.bc["scale"]))
REG.gamma.HC5     <- exp(REG.gamma.HC5.log)

MLE.gamma.HC1.log <- log(ssd_qgamma(.01, 
                           shape=fit.gamma$gamma$estimate["shape"],
                           scale=fit.gamma$gamma$estimate["scale"]))
MLE.gamma.HC1     <- exp(MLE.gamma.HC1.log)
REG.gamma.HC1.log <- log(ssd_qgamma(.01,
                           shape=bias.correct.gamma$mle.bc["shape"],
                           scale=bias.correct.gamma$mle.bc["scale"]))
REG.gamma.HC1     <- exp(REG.gamma.HC1.log)

    
```

The HC5 estimated from the MLE.gamma fit is `r round(MLE.gamma.HC5.log,2)` on the logarithmic concentrations
scale or `r round(MLE.gamma.HC5,3)` on the concentration scale.
The HC5 estimated after correcting for small sample bias is
`r round(REG.gamma.HC5.log,2)` on the logarithmic concentrations
scale or `r round(REG.gamma.HC5,3)` on the concentration scale. The ratio of these two HC5 values
is `r round(REG.gamma.HC5/MLE.gamma.HC5,3)`x on the concentration scale, 
i.e. the HCx based on the MLE is `r round(MLE.gamma.HC5/REG.gamma.HC5,1)`x larger on the concentration scale.

The differences between the HCx computed from the MLE and for
the bias corrected estimates will become more pronounced for small HCx values. For example,
the HC1 estimated from the MLE.gamma fit is `r round(MLE.gamma.HC1,3)` 
and the HC1 estimated using the biased corrected estimates
is `r formatC(REG.gamma.HC1,digits=6, format="f")` on the concentration scale. The ratio of these two values
is now `r round(REG.gamma.HC1/MLE.gamma.HC1,3)`x,
i.e. the HCx based on the MLE is `r round(MLE.gamma.HC1/REG.gamma.HC1,1)`x larger on the concentration scale.


We repeated a similar simulation study with the gamma distribution. The shape and scale
parameters were chosen to match the mean and variance of the log-normal distribution used
in the previous simulation study. 

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Simulation study to estimate effect of small sample bias on estimates of HC5 and HC1
# Use a gamma distribution with the same mean and sd as a log-normal (0,1) 

set.seed(23432)


# get the shape and scale parameter
mean.log <- 0
sd.log <- 1
mu <- exp(mean.log+.5*sd.log^2)
sd <- sqrt(exp(sd.log^2-1) * (exp(2*mean.log + sd.log^2)))
cat("Mean and variance of lognormal(0,1)", mu, sd, "\n")

# solve for shape and scale
scale = sd^2/mu
shape = mu/scale
cat("Estimated shape and scale ", shape, scale, "\n")
scale*shape
sqrt(shape*scale^2)

mean(rgamma(1000, shape=shape, scale=scale))
sd  (rgamma(1000, shape=shape, scale=scale))

sample.sizes <- c(6, 8, 10, 12, 15, 20, 25, 30, 50)

res <- plyr::ldply(sample.sizes, function(n, shape, scale, nsim=1000){
  fits <- plyr::ldply(1:nsim, function(sim, n, shape, scale){
     # generate sample of size n from a normal distribution with specified mean and variance
     #browser()
     sim.error=FALSE
     df <- data.frame(Conc= rgamma(n, shape=shape, scale=scale))
     
     # find the mle using ssd tools 
     fit <- ssd_fit_dists(df, dist="gamma")
     
     # find the bias corrected estimates
     gamma.pdf <- quote(1 /(scale ^ shape * gamma(shape)) * x ^ (shape - 1) * exp(-x / scale))

     # what is the log(density) functiong ingoring constants
     log.gamma.pdf <- quote(-shape * log(scale) - lgamma(shape) + shape * log(x) - x / scale)
     #browser()
     bias.correct <- try(coxsnell.bc(density = gamma.pdf, 
            logdensity = log.gamma.pdf, 
            n = n, 
            parms = c("shape", "scale"),
            mle = c(fit$gamma$estimate["shape"], 
                    fit$gamma$estimate["scale" ]), 
            lower = 0, upper = 'Inf'))
     if(class(bias.correct)=="try-error"){
        sim.error = TRUE
        cat("Gamma bias correct sim error ", n, sim, "\n")
        bias.correct <- NULL
        bias.correct$mle.bc <- c(shape=shape, scale=scale)
     }
   
     # find the hc5 and hc1 using the MLE and bias corrected values
     # compute the HC5 from the two fit
     actual.HC5.log <- log(ssd_qgamma(.05, shape=shape, scale=scale))
     actual.HC1.log <- log(ssd_qgamma(.01, shape=shape, scale=scale))
     actual.HC5     <- exp(actual.HC5.log)
     actual.HC1     <- exp(actual.HC1.log)
     MLE.HC5.log <- log(ssd_qgamma(.05,shape=fit$gamma$estimate["shape"], scale=fit$gamma$estimate["scale"]))
     MLE.HC5     <- exp(MLE.HC5.log)
     BC.HC5.log  <- log(ssd_qgamma(.05,shape = bias.correct$mle.bc["shape"], 
                                   scale = bias.correct$mle.bc["scale"] ))
     BC.HC5      <- exp(BC.HC5.log)

     MLE.HC1.log <- log(ssd_qgamma(.01, shape=fit$gamma$estimate["shape"], scale=  fit$gamma$estimate["scale"]))
     MLE.HC1     <- exp(MLE.HC1.log)
     BC.HC1.log  <- log(ssd_qgamma(.01,shape=bias.correct$mle.bc["shape"], 
                                   scale=bias.correct$mle.bc["scale"] ))
     BC.HC1      <- exp(BC.HC1.log)

     data.frame(n=n, sim=sim,  shape=shape, scale=scale,
              mle.shape = fit$gamma$estimate["shape"], 
              mle.scale =fit$gamma$estimate["scale"],
              bc.shape  = bias.correct$mle.bc["shape"],
              bc.scale  = bias.correct$mle.bc["scale"],
              actual.HC5.log,
              actual.HC5,
              actual.HC1.log,
              actual.HC1,
              MLE.HC5.log = MLE.HC5.log,
              MLE.HC5,
              BC.HC5.log,
              BC.HC5,
              MLE.HC1.log,
              MLE.HC1,
              BC.HC1.log,
              BC.HC1,
              sim.error=sim.error)
   }, n=n, shape=shape, scale=scale) 
   fits
}, shape=shape, scale=scale, nsim=nsim)

# remove simulation errors
sum(res$sim.error)
res <- res[ !res$sim.error, ]

head(res)
# summarize the output from the sishapelation
res.summary <- plyr::ddply(res,"n", plyr::summarize,
                           n=mean(n),
                           nsims=max(sim),
                           gamma.shape=mean(shape),
                           gamma.scale=mean(scale),
                           mean.mle.shape  = mean(mle.shape),
                           mean.mle.scale  = mean(mle.scale),
                           mean.bc.shape   = mean(bc.shape),
                           mean.bc.scale   = mean(bc.scale),
                           actual.HC5.log  = mean(actual.HC5.log),
                           actual.HC1.log  = mean(actual.HC1.log),
                           mean.mle.HC5.log= mean(MLE.HC5.log),
                           mean.mle.HC1.log= mean(MLE.HC1.log),
                           mean.bc.HC5.log = mean(BC.HC5.log),
                           mean.bc.HC1.log = mean(BC.HC1.log),
                           mean.mle.HC5    = mean(MLE.HC5),
                           mean.mle.HC1    = mean(MLE.HC1),
                           mean.bc.HC5     = mean(BC.HC5),
                           mean.bc.HC1     = mean(BC.HC1)
                           )
res.summary


plotdata <- reshape2::melt(res.summary,
                           id.vars="n",
                           value.name="value",
                           variable.name="Measure")
plotdata$Measure <- as.character(plotdata$Measure)
unique(plotdata$Measure)
str(plotdata)

plotdata$parameter <- car::recode(plotdata$Measure,
                          " 'mean.mle.shape'= 'Shape';
                            'mean.bc.shape' = 'Shape';
                            'mean.mle.scale'  ='Scale';
                            'mean.bc.scale'   ='Scale';
                            'mean.mle.HC5.log'='log(HC)';
                            'mean.mle.HC1.log'='log(HC)';
                            'mean.bc.HC5.log' ='log(HC)';
                            'mean.bc.HC1.log' ='log(HC)';
                            'mean.mle.HC5'    ='actual HC';
                            'mean.mle.HC1'    ='actual HC';
                            'mean.bc.HC5'     ='actual HC';
                            'mean.bc.HC1'     ='actual HC';

                          ")

plotdata$method <- car::recode(plotdata$Measure,
                           "'mean.mle.shape'  ='MLE';
                            'mean.bc.shape'   ='BC';
                            'mean.mle.scale'  ='MLE';
                            'mean.bc.scale'   ='BC';
                            'mean.mle.HC5.log'='MLE';
                            'mean.mle.HC1.log'='MLE';
                            'mean.bc.HC5.log' ='BC';
                            'mean.bc.HC1.log' ='BC';
                            'mean.mle.HC5'    ='MLE';
                            'mean.mle.HC1'    ='MLE';
                            'mean.bc.HC5'     ='BC';
                            'mean.bc.HC1'     ='BC';
                          ")

plotdata$HC <- car::recode(plotdata$Measure,
                           "'mean.mle.HC5.log'='HC5';
                            'mean.mle.HC1.log'='HC1';
                            'mean.bc.HC5.log' ='HC5';
                            'mean.bc.HC1.log' ='HC1';
                            'mean.mle.HC5'    ='HC5';
                            'mean.mle.HC1'    ='HC1';
                            'mean.bc.HC5'     ='HC5';
                            'mean.bc.HC1'     ='HC1';
                           else='NA';
                          ")

head(plotdata)

xtabs(~Measure+parameter,  data=plotdata, exclude=NULL, na.action=na.pass)
xtabs(~Measure+method   ,  data=plotdata, exclude=NULL, na.action=na.pass)
unique(plotdata$Measure)
select <- grepl("mle.shape$", plotdata$Measure) |
          grepl("mle.scale$", plotdata$Measure) |
          grepl("bc.shape$" , plotdata$Measure) |
          grepl("bc.scale$" , plotdata$Measure) |
          grepl("mle.HC5.log", plotdata$Measure) |
          grepl("mle.HC1.log", plotdata$Measure) |
          grepl("bc.HC5.log", plotdata$Measure) |
          grepl("bc.HC1.log", plotdata$Measure) |
          grepl("mle.HC5",    plotdata$Measure) |
          grepl("mle.HC1",    plotdata$Measure) |
          grepl("bc.HC5",     plotdata$Measure) |
          grepl("bc.HC1",     plotdata$Measure)

plotdata[select,]

true.parms <- data.frame(parameter=c("Shape",
                                     "Scale",
                                    "log(HC)",
                                    "log(HC)",
                                    "actual HC",
                                    'actual HC'),
                         value=c(shape,scale,
                                 log(ssd_qgamma(.05, shape=shape, scale=scale)), log(ssd_qgamma(.01, shape=shape, scale=scale)), 
                                 ssd_qgamma(.05, shape=shape, scale=scale), ssd_qgamma(.01, shape=shape, scale=scale)),
                         linetype=c("solid","solid","dashed","solid","dashed","solid"), stringsAsFactors=FALSE)
true.parms

sim.plot <- ggplot(data=plotdata[select,], aes(x=n, y=value,color=method, linetype=HC))+
  ggtitle("Performance of mle and bias corrected estimators",
           subtitle=paste("Gamma distribution with shape= ", round(shape,2), ' and scale =',round(scale,2),sep=""))+
  geom_point(position=position_jitter(w=0.4))+
  geom_line( position=position_jitter(w=0.4))+
  facet_wrap(~parameter, ncol=2,scales="free")+
  geom_hline(data=true.parms, aes(yintercept=value), linetype=true.parms$linetype)+
  xlab("Sample size")+
  scale_linetype_discrete(na.value="solid", breaks=c("HC1","HC5") )

```
```{r echo=FALSE, warning=FALSE, message=FALSE}
sim.plot
```

The MLEs are biased in small-samples for both the shape and scale (bottom row of plots) but the
small-sample bias declines as sample size increases (as expected). The biases of the
two parameters are in opposite directions (i.e. one bias is positive and one bias is negative).
The bias corrected estimates
are unbiased (as expected). 

The estimated $\log{HC}$ (upper right plot) based on the MLE is slightly biased upwards (i.e. larger) than the true
values but the bias rapidly declines with sample size (as expected). Rather surprisingly, the
estimated HC5 and HC1 values using the bias-corrected estimates are biased downwards, likely an artefact
of the non-linear transformation from scale and shape to the HCx.

Finally, the estimated $HC1$ and $HC5$ values are again biased upwards (upper left plot) based
on the MLEs, but the estimated $HCx$ values based on the bias-corrected estimates appear to exhibit 
less bias despite the bias in the $\log(HCx)$ values.

## Recommendations

In cases with reasonably large sample sizes (around 15+), the small sample bias is 
unlikely to be of concern given the uncertainty in the endpoints actually used
for the fit, and the uncertainty generated for the HCx from the model averaging process.

The small sample bias in the estimates is expected to affect the smaller 
HCx values (e.g. HC1) more than larger HCx values (e.g. HC5). 
This is not unexpected because you are
trying to extrapolate out to the extreme tails of the distribution where there
is typically no data available and small changes to parameter values can have
large impacts on the extreme tails.

For smaller sample sizes, a similar exercise as above can be used to 
estimate the impact of the small sample bias. Howeverr, for small sample sizes,
this exercise may be akin to "fiddling while Rome burns", i.e, this does not change the basic
problems with small sample sizes including (a) most distributions will have adequate fits and it is unlikely be possible to discriminate between distributions;
and (b) extrapoloting even to a moderate tail fraction (e.g. HC5) is very, very
dependent on the chosen distribution; (c) there is no data available
to support even moderate extrapolation to tail proportions. Higher certainty in 
the estimates can only be obtained by increasing sample sizes.

```{r, results = "asis", echo = FALSE}
cat(licensing_md())
```
